---
title: "Toronto Rental Housing"
author: Lee Doucet
date: January 31st, 2021
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("opendatatoronto")
# install.packages('knitr', dependencies = TRUE)
# install.packages("kableExtra")
#install.packages("bibtex")
library(bibtex)
library(opendatatoronto)
library(tidyverse)
#library(knitr)
library(kableExtra)

```
 
```{r Getting Data from Open Data Toronto, include=FALSE}

rawApartmentEval <- 
  opendatatoronto::search_packages("Apartment Building Evaluation") %>%
  opendatatoronto::list_package_resources() %>%
  filter(name == "Apartment Building Evaluation") %>% # This is the row that we are interested in
  select(id) %>%
  opendatatoronto::get_resource()

## Apartment Regulation
rawApartmentReg <- 
  opendatatoronto::search_packages("Building Registration") %>%
  opendatatoronto::list_package_resources() %>%
  filter(name == "Apartment Building Registration Data") %>% #This is the row that we are interested in
  select(id) %>%
  opendatatoronto::get_resource()

```

# Introduction 

	The City of Toronto protects renters in apartment buildings through a program 
called RentSafeTo where they enact a bylaw enforcement program, Apartment 
Building Standards (ABS). Specific maintenance standards and procedures are 
laid out for the tenant’s behalf, including their right to have a landlord 
respond to urgent service requests (e.g. no heat or water) within 24 hours of
being reported. Tenants who experience problems with these standards are able 
to contact the RentSafeTo team through 311 to escalate their concerns. An 
important consideration is these protections apply to apartment buildings
alone. Whose definition is that an “apartment building is a purpose-built 
rental building with 3 more storeys and 10 or more rental units but does not
include long-term care, a licensed retirement home or a housing 
co-operative”[@citeRentSafeTo]. This also excludes condo buildings,
townhomes, and units residing within the basement level or main floor of a 
private home from the program. 
	
	With Toronto’s growing population and concerns over equitable access to 
housing, a one-size fits all approach with strictly private rentals would 
not address the housing concerns of all residents within Toronto. To 
ensure everyone has equal access to adequate housing, the Ontario Human 
Rights Commission (OHRC) promotes the increased availability of social 
housing in compliance with Canada’s commitment to provide adequate housing
for all Canadians [@citeOHRC]. The largest social housing provider in 
Canada is the Toronto Community Housing Corporation (TCHC) and manages
60,000 rental housing units with a mandate to “to provide clean, safe, 
secure homes in a state of good repair to low and mode and moderate-income 
households, including seniors, families, singles, refugees, recent immigrants
to Canada and people with special needs” [@citeTCHC]. 
	
	The focus of this paper will be an exploratory analysis of the apartment 
buildings in Toronto, examining their state of condition and much progression
Canada has made in regards to their policy goal of providing adequate housing
for all Canadians. To answer that question, data will be pulled from the
Toronto Open Data Portal exclusively. Once that has been investigated, the
research questions will shift to examine different facets of the Toronto
rental market that may have the opportunity to provide important insights 
to social housing advocates such 

```{r Combining Data Sets & Cleaning up, include=FALSE}

apartmentReg <- rawApartmentReg
apartmentEval <- rawApartmentEval

apartmentReg <-
    apartmentReg %>%        #Removing unessential columns
    select(RSN, PCODE,APPROVED_FIRE_SAFETY_PLAN, FIRE_ALARM,
            IS_THERE_EMERGENCY_POWER, NO_BARRIER_FREE_ACCESSBLE_UNITS,
            PROP_MANAGEMENT_COMPANY_NAME)

apartmentEval <- 
     apartmentEval %>%
     select(RSN,CONFIRMED_UNITS, PROPERTY_TYPE, YEAR_BUILT,
            SECURITY,CONFIRMED_UNITS, WARD, SCORE)

allData <- merge(apartmentReg,apartmentEval, by = "RSN") #Common identifier in both data sets 

allData$SCORE <- as.numeric(allData$SCORE) # Making Apartment Score numeric

allData <- allData[!is.na(allData$SCORE),] #Removing 13 NA's 

allData <-       #Cleaning up names too long and full of CAPS
  rename(allData, c("pCode" = "PCODE", "firePlan" = "APPROVED_FIRE_SAFETY_PLAN", 
                   "fireAlarm" = "FIRE_ALARM", "rentalScore" = "SCORE",
                   "emergPower" = "IS_THERE_EMERGENCY_POWER",
                   "rentalUnits" = "CONFIRMED_UNITS",
                   "propertyManager" = "PROP_MANAGEMENT_COMPANY_NAME",
                   "propertyType" = "PROPERTY_TYPE", "yearBuilt" = "YEAR_BUILT",
                   "security" = "SECURITY", "ward" = "WARD", "rsn" = "RSN",
                   "accessUnits" ="NO_BARRIER_FREE_ACCESSBLE_UNITS" ))

``` 

#Data 
	
##Methodology 
	
	  To investigate the rental apartment housing in Toronto, their open data 
portal was accessed from theirwebsite and a search was conducted for 
apartment.The query provided two results, Apartment Building Evaluation with 
a gold data quality score , and Apartment Building Registration with a silver 
data quality score. Data sets are published by Municipal Licensing & Standards 
and covered under an Open Government License – Toronto, for distribution. 
	
## Key Features 
	
	The main benefit of using these two data sets is their interoperability with 
each other through the specific building identification number (RSN). This
allows the combination of the regulation attributes (e.g. Fire Code) and the
evaluation metrics (e.g. Building Score) to test if there are any 
correlations in the data. With the data conducted by inspectors there is 
strong reason to have confidence in the validity of what data has been 
collected. There is also the social good that comes from this data as people
can gain more insight into what buildings perform better and which ones 
struggle with compliance. Having clear metrics like Score allows for various 
other attributes to be tested against it to look for variation in the data, 
with our first test being if the type of property has any impact on the 
overall score of the building. The items that were included in Score are 
also broken down so the user can test different attributes (e.g. Security) 
if they want a narrower focus in their research. 

	Within the data there is a range of the completeness with fields such as 
Amenities Available having most of displayed NA. Other parts of the data 
are contradictory as 2362 Islington Ave lists no pets but, in another field,
says only small dogs. Many columns require extensive cleaning for use, 
Property Management Company for instance has many variations of spellings 
which make use difficult. There are data fields that are not present that 
would be very helpful such as average rent or number of validated violations 
reported against a building. All apartments that do not fall within the
definition are also missing and add a significant amount to the rental
inventory. It’s also not clear what fields account for what percentage of 
the total score if certain fields contain more weight or not. 

	
	From an ethics perspective, the primary concern is between whom the 
relationship is between in regards to the data. In each data set, it’s 
strictly how the building performs based on observable metrics but does not 
consider the lifeblood of the building, the residents. If a building gets a 
security score of 4 out of 5, for whom is that score based on, a male or a 
female or even a visible minority? A woman walking from her car at night 
would have much different needs than a woman late at night. The evaluation 
also only takes place within the building common areas so there is indication 
of what is transpiring inside tenant space. Landlords be would incentivized 
to do the minimal inside tenant space to avoid interactions ABS complaints 
and then too maximize the common areas of the building to attract new tenants
and receive high evaluation scores. 

##Merging the Data Sets 
	
	To make the data easier to work with, Apartment Evaluations was combined 
with Regulations by the mutual column RSN. With 100 columns together between 
the two data sets, 13 were selected based on their usefulness for a direct 
comparison (e.g. Score, Property Type, Postal Code) and other being 
attributes to investigate (e.g. Emergency Power, Fire Alarm status). Once 
they were organized together, there were deemed useable for analysis. 

```{r Data Prep for Chart 1 - Rental Summary / Frequency Table + Averages, include=FALSE }

chart1 <-
  allData %>%
  select(propertyType, rentalScore) #Want to compare these two columns 

property1 <-  #Average Rental Score of each property type 
  aggregate(. ~ propertyType, chart1, mean) 

property2 <- aggregate(cbind(count = propertyType) ~ propertyType, 
                  data = allData, 
                  FUN = function(x){NROW(x)}) # Sum Property Type

chart1 <- merge(property1,property2, by = "propertyType")

chart1 <- #Getting Proper Names 
  rename(chart1, c("Property Type" = "propertyType", 
                   "Average Rental Score" = "rentalScore",
                   "Total Amount" = "count"))

```

## Research Question 1 – Do total scores differ amongst the three different 
##rental building types? 


```{r Summary Chart , echo=FALSE}

chart1 %>% #Creating a chart to display the data for Property Type by Score/Count
  knitr::kable(caption = "Table 1: Summary statistics", digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width =  FALSE)

```

#Graph 1 

  Our data is from [@citeR]

```{r Graphical Version of Chart 1, echo=FALSE}

graph1 <- 
ggplot(data = allData, aes(x = propertyType, y = rentalScore, fill = propertyType)) +
  geom_boxplot() +
  ylab("Building Evaluation (%)") +
  xlab("Property Type")
  ggtitle("                 Distribution Scores Between Apartment Types ")
  
graph1

```
## Research Question 2 - Does Postal Code Impact Rental Score? 

```{r Graph 2 Data, echo = FALSE }

graph2Data <-
  allData %>%
  select(pCode,rentalScore) #Gathering data for the graph

graph2Avg <-
  aggregate(. ~ pCode,graph2Data, mean) #Getting Rental Score Average

graph2Count <- aggregate(cbind(count = pCode) ~ pCode, 
                   data = graph2Data, 
                   FUN = function(x){NROW(x)}) # Building Count Per Postal Code

graph2Combine <- merge(graph2Avg,graph2Count, by = "pCode") #Combination

graph2 <- 
attach(graph2Combine)
plot(count, rentalScore, main="Relationship Between Average Rental Price and Postal Code",
     xlab="Number of Rentals in Postal Code ", ylab="Score Per Rental ", pch=19) +
     abline(lm(rentalScore~count), col="red") 
graph2


```

```{r Chart 2 Data, echo = FALSE }

chart2Data <-
  allData %>%
  select(propertyType,rentalUnits, accessUnits)
     
chart2Data <-
  chart2Data %>%
  filter(rentalUnits < 750)

chart2Data$accessUnits <- as.numeric(chart2Data$accessUnits)
  
chart2Agg <-  
    aggregate(. ~ propertyType, chart2Data, sum) 

chart2 <-
mutate(chart2Agg, Percentage = rentalUnits / accessUnits)

chart2 <-
  rename(chart2, c("Property Type" = "propertyType",
                   "Rental Units" = "rentalUnits",
                   "Accessibility Units" = "accessUnits"))

chart2 %>% #Creating a chart to display the data for Property Type by Score/Count
  knitr::kable(caption = "Table 2: Summary statistics", digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width =  FALSE)


```

##References 


```{r Scatter Graph, echo= FALSE}






```


