---
title: "Private and Government Rental Apartment Compairson in Toronto"
author: "Lee Doucet"
date:  "February 1st 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("opendatatoronto")
# install.packages('knitr', dependencies = TRUE)
# install.packages("kableExtra")
# install.packages("bibtex")
# install.packages("here")
library(here)
library(bibtex)
library(opendatatoronto)
library(tidyverse)
library(kableExtra)

```
# Abstract 

Research that uses Toronto Open Data to compare the differences in 
the Toronto apartment rental markets between the private sector and social 
housing offered by the government. Brings together the data sets from apartment
evaluations and regulations conducted by city inspectors and cleans them up 
for analysis.Finally, builds upon the first research question and investigates
differences in accessibility between private and social housing to determine
what level of access some of Toronto's most vulnerable have afforded to them. It 
was written and analyzed using [@citeR] and the packages available through open
software use. 



```{r Getting Data from Open Data Toronto, include=FALSE}

rawApartmentEval <- 
  opendatatoronto::search_packages("Apartment Building Evaluation") %>%
  opendatatoronto::list_package_resources() %>%
  filter(name == "Apartment Building Evaluation") %>% # This is the row that we are interested in
  select(id) %>%
  opendatatoronto::get_resource()

## Apartment Regulation
rawApartmentReg <- 
  opendatatoronto::search_packages("Building Registration") %>%
  opendatatoronto::list_package_resources() %>%
  filter(name == "Apartment Building Registration Data") %>% #This is the row that we are interested in
  select(id) %>%
  opendatatoronto::get_resource()

write.csv(here::here( "inputs/data/toronto_housing_data.csv"))
```

# Introduction 

The City of Toronto protects renters in apartment buildings through a program 
called RentSafeTo where they enact a bylaw enforcement program, Apartment 
Building Standards (ABS). Specific maintenance standards and procedures are 
laid out for the tenant’s behalf, including their right to have a landlord 
respond to urgent service requests (e.g. no heat or water) within 24 hours of
being reported. Tenants who experience problems with these standards are able 
to contact the RentSafeTo team through 311 to escalate their concerns. An 
important consideration is these protections apply to apartment buildings
alone. Whose definition is that an “apartment building is a purpose-built 
rental building with 3 more storeys and 10 or more rental units but does not
include long-term care, a licensed retirement home or a housing 
co-operative”[@citeRentSafe]. This also excludes condo buildings,
townhomes, and units residing within the basement level or main floor of a 
private home from the program. 

	
With Toronto’s growing population and concerns over equitable access to 
housing, a one-size fits all approach with strictly private rentals would 
not address the housing concerns of all residents within Toronto. To 
ensure everyone has equal access to adequate housing, the Ontario Human 
Rights Commission (OHRC) promotes the increased availability of social 
housing in compliance with Canada’s commitment to provide adequate housing
for all Canadians [@citeOHRC]. The largest social housing provider in 
Canada is the Toronto Community Housing Corporation (TCHC) and manages
60,000 rental housing units with a mandate to “to provide clean, safe, 
secure homes in a state of good repair to low and mode and moderate-income 
households, including seniors, families, singles, refugees, recent immigrants
to Canada and people with special needs” [@citeTCHC]. 

	
The focus of this paper will be an exploratory analysis of the apartment 
buildings in Toronto, examining their state of condition and much progression
Canada has made in regards to their policy goal of providing adequate housing
for all Canadians. To answer that question, data will be pulled from the
Toronto Open Data Portal exclusively. Once that has been investigated, any 
other research insights that present themselves will be examined that 
pertains to the Toronto rental market. The goal is provide a framework of data 
that can be built upon for policy planners at the city or advocacy organizations
such as ACORN Toronto.Finally, as this research is exploratory, it will hope to 
highlights future areas of research opportunities and suggest new data avenues 
for the Toronto Open Data Portal to collect. 

```{r Combining Data Sets & Cleaning up, include=FALSE}

apartmentReg <- rawApartmentReg
apartmentEval <- rawApartmentEval

apartmentReg <-
    apartmentReg %>%        #Removing unessential columns
    select(RSN, PCODE,APPROVED_FIRE_SAFETY_PLAN, FIRE_ALARM,
            IS_THERE_EMERGENCY_POWER, NO_BARRIER_FREE_ACCESSBLE_UNITS,
            PROP_MANAGEMENT_COMPANY_NAME)

apartmentEval <- 
     apartmentEval %>%
     select(RSN,CONFIRMED_UNITS, PROPERTY_TYPE, YEAR_BUILT,
            SECURITY,CONFIRMED_UNITS, WARD, SCORE)

allData <- merge(apartmentReg,apartmentEval, by = "RSN") #Common identifier in both data sets 

allData$SCORE <- as.numeric(allData$SCORE) # Making Apartment Score numeric

allData <- allData[!is.na(allData$SCORE),] #Removing 13 NA's 

allData <-       #Cleaning up names too long and full of CAPS
  rename(allData, c("pCode" = "PCODE", "firePlan" = "APPROVED_FIRE_SAFETY_PLAN", 
                   "fireAlarm" = "FIRE_ALARM", "rentalScore" = "SCORE",
                   "emergPower" = "IS_THERE_EMERGENCY_POWER",
                   "rentalUnits" = "CONFIRMED_UNITS",
                   "propertyManager" = "PROP_MANAGEMENT_COMPANY_NAME",
                   "propertyType" = "PROPERTY_TYPE", "yearBuilt" = "YEAR_BUILT",
                   "security" = "SECURITY", "ward" = "WARD", "rsn" = "RSN",
                   "accessUnits" ="NO_BARRIER_FREE_ACCESSBLE_UNITS" ))

``` 

# Data 


## Methodology 
	
	
To investigate the rental apartment housing in Toronto, their open data 
portal was accessed from their website where a search was conducted for data 
sets related to apartments The query provided two results, Apartment Building 
Evaluation with a gold data quality score and Apartment Building Registration 
with a silver data quality score. Data sets are published by Municipal 
Licensing & Standards and covered under an Open Government License – Toronto, 
for distribution. 
	

## Data Description 


Both data sets include all apartment rental buildings in Toronto as per the 
aforementioned definition in the introduction. Building owners are required to 
register and renew annually within the compliance of the RentSafeTO program.
Audits are conducted by the city inspectors at least once every three years with 
increased frequency of inspections depending on the amount of property 
standard issues currently present in the building.The program has been in 
effect since 2017 when the first buildings were registered.
	
# Data Sets 
	
#### Apartment Building Evaluation Data Set Highlights [@citeEval]

- Total Data: 3462 rows of data across 70 variables	
-	RSN: A 7-digit number to Identify the building across different data sets 
-	Property Type: Either Private, Social Housing, or TCHC
-	Ward: Which of the 25 wards in Toronto does this building reside 
-	Score: Out of 100 what did this building score 
-	Confirmed Units: Total amount of units present in the building 
-	Year Built: When was this building built 
- Security: Gives the security score (out of five)

#### Apartment Building Registration Data Set Highlights [@citeReg]

- Total Data: 3461 rows of data across 33 variables 
-	RSN: A 7-digit number to Identify the building across different data sets 
-	Property Type: Either Private, Social Housing, or TCHC
-	PCODE: Postal Code of the building 
-	Property Management Company Name: Identify the owner of the building
-	No Barrier Free Accessible Units: Number of accessible units present in 
  the building
- Fir Alarm: Does the building have an active fire alarm
- Approved Fire Safety Plan: Does the building have an approved fire safety plan


\newpage
	
## Key Features 
	
	
The main benefit of using these two data sets is their interoperability with 
each other through the specific building identification number (RSN). This
allows the combination of the regulation attributes (e.g. Fire Code) and the
evaluation metrics (e.g. Building Score) to test if there are any 
correlations in the data. With the data conducted by inspectors there is 
strong reason to have confidence in the validity of what data has been 
collected. There is also the greater social good that comes from open data 
that informs citizens which buildings perform the best and which ones struggle
with compliance. Having a clear metrics such as Building Score allows for 
other attributes to be tested against it to look for relationships, or the lack
of them in the data. For example, we can examine the relationship between the
property type of the building while analyzing it against the overall score of
the building. The metric Score can also be further broken down into any 
quantifiable attributes presented (e.g. Security) that a part of the overall 
score if a narrower focus is desired for research. 

Within the data there is a range of the completeness among the data fields. For
instance, Amenities Available has most of its data displayed as NA. Some fields
hold contradictory information with related fields with 2362 Islington Ave 
listing no pets, but in another field, says that small dogs are okay. 
Many fields require extensive cleaning for use, the Property Management Company 
field has many variations of spellings or additional punctuation that makes 
use difficult without proper cleaning of the data. 
 
There are also data fields that are not present that would have been very 
helpful such as average rent or total number of validated violations 
reported against a building. All apartments that do not fall within the
definition are also missing and add a significant amount to the rental
inventory. It’s also not clear what fields account for what percentage of 
the total score if certain fields contain more weight or not. For instance,
not all buildings have balconies, does that mean that the field is not 
calculated in total score for all buildings or just for buildings without a 
balcony.

From an ethical perspective, the primary concern is between whom the 
relationship is between in regards to the data. In each data set, it’s 
strictly how the building performs based on observable metrics but does not 
consider the lifeblood of the building, the residents. If a building gets a 
security score of 4 out of 5, for whom is that score based on, a male or a 
female, or even a visible minority? A woman walking from her car at night 
would have much different security needs to feel safe at night than a man. The 
The evaluation also only takes place within the building common areas so there 
is little indication of what is transpiring inside tenant space. The data does
not say the tenants appliances are being taken care by the landlords and tenant.
Landlords could possibly be incentivized to do the least amount of upkeep inside 
the tenant space to prevent the tenant from lodging a complaint. This could 
actually be happening and there would be no way to tell from this data. This is 
why it is important to use open data as a starting point to investigate and 
ask better questions that are informed with the data. 




## Merging the Data Sets 

	
To make the data easier to work with, Apartment Evaluations was combined 
with Regulations by the mutual column RSN. With 100 columns together between 
the two data sets, 13 were selected based on their usefulness for a direct 
comparison (e.g. Score, Property Type, Postal Code) and other being 
attributes to investigate (e.g. Emergency Power, Fire Alarm status). Once 
they were organized together, the data was examined and areas for research
analysis were investigated. 

\newpage

```{r Data Prep for Chart 1 - Rental Summary / Frequency Table + Averages, include=FALSE }

chart1 <-
  allData %>%
  select(propertyType, rentalScore) #Want to compare these two columns 

property1 <-  #Average Rental Score of each property type 
  aggregate(. ~ propertyType, chart1, mean) 

property2 <- aggregate(cbind(count = propertyType) ~ propertyType, 
                  data = allData, 
                  FUN = function(x){NROW(x)}) # Sum Property Type

chart1 <- merge(property1,property2, by = "propertyType")

chart1 <- #Getting Proper Names 
  rename(chart1, c("Property Type" = "propertyType", 
                   "Average Rental Score" = "rentalScore",
                   "Total Amount" = "count"))

```


# Research Question - Does the Rental Score Vary Across Property Types?



The results from Graph 1 show how the data is distributed among the three 
property types. Each box plot shows the range of the data where most of it 
located with the black line representing the median of the data. From this
graph, social housing appears to have scored the highest with private rentals 
very close. TCHC housing is lagging behind the both of them with much lower 
scores. It also tells us the outliers from the data, demonstrating that private
rental housing has the best overall units but also the worst. An abundance of 
lower units on the range for private rentals is not surprising considering 
some of the strategies enacted through the financialization of the private 
rental housing market. Take for example in the non-gentrifying parts of New 
York’s outer boroughs, the private equity firms that own them use strategic u
nder maintenance for those who are low-income and immigrant populations. 
This group has little options to find better lodgings so they must suffer 
reduced services and higher rents[@citeAUG]. Let’s examine the actual numbers 
of each property type present. 

$$\\[1in]$$


```{r Summary Chart , echo=FALSE,  fig.pos="H"}

graph1 <- 
ggplot(data = allData, aes(x = propertyType, y = rentalScore, fill = propertyType)) +
  geom_boxplot() +
  ylab("Building Evaluation (%)") +
  xlab("Property Type") +
  ggtitle("Graph 1 - Distribution Scores Between Apartment Types ") +
  labs(caption = "(Based on data from ggPlot2 (Wickham 2016) & R (R Core Team 2020))") +
  scale_fill_discrete(name = "Property Type")
graph1


```



\newpage


```{r Graphical Version of Chart 1, echo=FALSE,  fig.pos="H"}

chart1 %>% #Creating a chart to display the data for Property Type by Score/Count
  knitr::kable(caption = "Summary statistics", digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width =  FALSE)


```

Table 1 was generated by Kable[@citeKABLE] & R [@citeR]


Table 1 makes it clearer that the social housing and TCHC are vastly under 
presented in the total amount of rental apartments in Toronto. This makes 
comparison more challenging to compare them both directly and accurately when
there is such a discrepancy. This does provides a new direction with examining 
why there is such a difference in the total number of rental units built in the 
first place. 

$$\\[1in]$$

```{r Line Graph, echo= FALSE,  fig.pos="H"}

builtGraph <- 
  allData %>%
  select(propertyType,yearBuilt) 

sortBuiltGraph <-
  builtGraph %>%
  filter(builtGraph$yearBuilt > 1966) #range picked to make the data look more clear

sortBuiltGraph <-
  filter(sortBuiltGraph, propertyType == "SOCIAL HOUSING") 

builtGraph <-
  aggregate(. ~ yearBuilt,sortBuiltGraph, length)

  
ggplot(data=builtGraph, aes(x=yearBuilt, propertyType, group=1)) +
  geom_step(color="blue")+
  geom_point() +
  xlab("Year Built") +
  ylab("Number of Social Housing Units Built") +
  scale_color_grey() + theme_classic() +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("                    Graph 2 - Decline of Social Housing Being Built") +
  labs(caption = "(Based on data from ggPlot2 (Wickham 2016) & R (R Core Team 2020))") 
  
  


builtGraph2 <- 
  allData %>%
  select(propertyType,yearBuilt) 

sortBuiltGraph2 <-
  builtGraph2 %>%
  filter(builtGraph2$yearBuilt > 1966) #range picked to make the data look more clear

sortBuiltGraph2 <-
  filter(sortBuiltGraph2, propertyType == "TCHC") 

builtGraph2 <-
  aggregate(. ~ yearBuilt,sortBuiltGraph2, length)

  
ggplot(data=builtGraph2, aes(x=yearBuilt, propertyType, group=1)) +
  geom_step(color="red")+
  geom_point() +
  xlab("Year Built") +
  ylab(" Number of TCHC Rental Apartments Built") +
  scale_color_grey() + theme_classic() +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("                      Graph 3 - Decline of TCHC Rentals Being Built ") +
labs(caption = "(Based on data from ggPlot2 (Wickham 2016) & R (R Core Team 2020))") 
  
```

$$\\[1in]$$

Looking at Graph 3, it shows the gradual decline of TCHC apartments being built
since the 1960’s with a sharp drop off in the early 1990’s. With Graph 2, the 
Social Housing shows lower builds per year with a sharp increase in the early 1990’s, 
followed by a steady drop off in supply. Both these line graphs are consistent
with the Canadian government eliminating funding for new social housing in 1993
and downloading that responsibility to the provincial governments whom in 1995. 
After that a Conservative government was elected and they further downloaded 
these responsibilities to the municipal governments [@citeAUG]. There has been 
little help from the private sector to fill this void in Toronto that has 
shifted their development away from rental housing apartments towards the more 
higher profits from condominium development [@citeAUG]. 

\newpage

Table 2 was generated by Kable [@citeKABLE] & R [@citeR]

# Research Question 2 – Does Access to Accessible Units Change Based on Property


The research presented has presented the declining volume of rental 
apartments and referenced tactics displayed by New York by private equity firms
that try to gain as much profit out of low-income tenants while offering less 
than exemplary experiences. Using our data, we can try and gain some insights 
into the apartment rental offerings to those requiring accessibility services. 



```{r Chart 2 Data, echo = FALSE,  fig.pos="H"}

chart2Data <-
  allData %>%
  select(propertyType,rentalUnits, accessUnits)
     
chart2Data <-
  chart2Data %>%
  filter(rentalUnits < 750)

chart2Data$accessUnits <- as.numeric(chart2Data$accessUnits)
chart2Data$rentalUnits <- as.numeric(chart2Data$rentalUnits)
  
chart2Agg <-  
    aggregate(. ~ propertyType, chart2Data, sum) 

chart2 <-
mutate(chart2Agg, Percentage = accessUnits / rentalUnits * 100)

chart2 <-
  rename(chart2, c("Property Type" = "propertyType",
                   "Rental Units" = "rentalUnits",
                   "Accessibility Units" = "accessUnits"))

chart2 %>% 
  knitr::kable(caption = "Summary statistics", digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width =  FALSE)
```



From Table 2 we can gather a snapshot of how each property type scores on 
the number of barrier free accessible units. The private sector is doing the 
best at 11.94%, followed closely by Social Housing. What is really unfortunate
with the data is how poor the TCHC is doing with less than 2% of units being 
barrier free accessible. This increases the challenge to find appropriate 
housing for those with accessibility concerns that also rely on 
government-funded subsidized housing as they cannot compete in the private 
market. Taking into consideration the already long waits for access to Toronto 
Community Housing which at present is between seven to eight years long 
[@citeOHRC]. Although the numbers are better in the private sector, there are 
concerns over discrimination. In a study in Sweden, rental applicants who 
inquired about a unit’s handicap accessibility when they contacted a prospective
landlord received statistically fewer responses and less positive responses 
than the control group that did not mention the disability[@citeMOM]. 
Meaning that although the private sectormay provide better accessibility, it
does not account for any struggle perceived or real, from tenants trying to 
obtain appropriate housing. 



\newpage

# Conclusion 

Toronto providing their open data is a boon for researchers looking for rental 
housing data and others curious about the current state of the market. An 
increase in access to this data can hopefully lead to better public policy 
or at the very minimum, better civic engagement in housing debates which 
access to is a international commitment by Canada. Within the context of the 
open data there is a connection between Toronto and the larger global 
phenomenon of governments not providing adequate numbers of social geared 
housing since the 1990’s. This is concerning as the private sector has little
incentive to address this as they are focused on the higher profitability of 
condominiums in the city. Even more concerning is the state of the TCHC housing,
as it scores lower than the private sector while being the refuge for Toronto’s 
most vulnerable. It is the belief of the author that governments must provide 
additional funding to bolster the supply of units and take more responsibility 
in the active care of these units. Considering it takes upwards of 2-3 years 
to build an apartment building when the first shovel hits the ground, 
the government must move forward on these actions immediately in hopes for a 
timely solution. It is the hope of this paper that further analysis can be 
conducted through the release of more open data and the work built upon 
by whomever has the desire and will. 


```{r Save The Data, echo= FALSE}

```


# References 
